Laves polyhedra in synthetic tennantite, Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}, and its lattice dynamics

### 2021-01-26 paper_tennantite.tex
comparison added

The tennantite-tetrahedri\textcolor{red}{t}e group represents a large class of inorganic compounds including complex copper sulfosalts with the general formula Cu\textsubscript{10+x}M\textsubscript{2-x}X\textsubscript{4}S\textsubscript{13}, where M is a transition metal (Zn, Fe, Hg, etc.), and X is a pnictogen (As, Sb, or Bi) \cite{Makovicky_2006}, with the pure tennantite itself represented by the formula Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}.
Despite being known for over two centuries, these minerals are still under investigation, both due to  structural peculiarities and potential applications. Arsenic- and antimony-containing compounds of this class are the advanced thermoelectric materials with a low lattice thermal conductivity\cite{Sootsman2009,Chetty2015}.

\textcolor{red}{The tennantite structure treated as sphalerite-type and belongs to \hmn{I -4 3 m} space group\cite{yaroslavzev2019,Makovicky_2006}.
There are three copper position: Cu1 is tetrahedrally coordinated by four sulpfur atoms, Cu2 is coordinated by three sulpfur atoms and Cu21 is tetrahedrally coordinated by three sulpfur and arsenic atoms.
Cu21 and Cu2 atoms possess very large atomic displacements perpendicular to the plane of the triangle (Fig.~\ref{fig:generel_view} right).}
In a case of tennantite- and tetrahedrite-based materials, it is generally considered, that a low lattice thermal conductivity is caused by hybridized Cu {\it 3d}  and chalcogen \textcolor{red}{{\it p}} orbitals\cite{Lai2015}.
Such a hybridization leads to a bonding asymmetry, and out-of-plane low frequency rattling modes, which are quasi-localized and anharmonic\cite{May2016,Bouyrie2015}.

@@

Some authors note\cite{DiBenedetto2005,Tanaka2015}, that asymmetric bonds give rise to magnetic properties and, in some cases, to antiferromagnetic (AFM) ordering\cite{yaroslavzev2019,Blandy_2018}.
In both kinds of materials, deviations from the Curie-Weiss law in the temperature dependencies of the magnetization were revealed: at 124~K\cite{Hathwar2019} in tennantite, and at 84~K\cite{Nasonova2016_2} in tetrahedrite.

@@

Large static values for Cu2 atom (Table~\ref{tab:enst_temp}) indicate the potential instability of this position. Even after dividing Cu2 position into Cu2 and Cu21, the Cu2 position remains highly disordered. This disordering is reliably described by the anharmonic model of atomic displacements using the Gram-Charlier expansion to the 4th order.

@@

Those modes were observed in tetrahedrites\cite{May2016}, and it was noted that they were caused by strongly anharmonic vibrations of the trigonally-bonded Cu atoms.
Within this approach, the Einstein solid model is used to describe independent harmonic oscillators.
If so, the low energy spectrum should consist of peaks with the Einstein's energies, which we found by DebyeFit (Table~\ref{tab:enst_temp}).
Results of calculations are presented in Figure~\ref{fig:low_energ_raman}.
The Einstein's energies from structural study were used as the initial points for the fit, and in Table~\ref{tab:ram_fit} the initial  and fitted wavenumbers are presented.
The low energy $l_{1}$ and $l_{2}$ peaks are well described by Einstein's energies of Cu2, Cu21, Cu1, and As (66, 40, 79, and 122 cm\textsuperscript{-1} respectively).
Thus, the reasons for the strong anharmonic behavior, described in the papers, is a dynamic disorder of the copper atoms. Moreover, arsenic atoms are shifted by the copper atoms.
Consequently, this leads to out-of-plane low frequency rattling modes, and a low lattice thermal conductivity.

@@

\section{Conclusions}\label{sec:level1}
In conclusion, the structural variety of synthetic tennantite, Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13},  leads to  a dynamic disorder of the copper atoms, asymmetric bonds, and inconsistent electronic densities.
This brings to the out-of-plane low frequency rattling modes, well described by Einstein characteristic temperatures: 74, 104, 115, and 185~K.
The peculiarities of the tennantite crystal and electronic structures lead to asymmetric Fermi surface, thus makinkg tennantite--tetrahedrite group compounds attractive research objects.

>>>>

The tennantite-tetrahedri\textcolor{red}{t}e group represents a large class of inorganic compounds including complex copper sulfosalts with the general formula Cu\textsubscript{10+x}M\textsubscript{2-x}X\textsubscript{4}S\textsubscript{13}, where M is a transition metal (Zn, Fe, Hg, etc.), and X is a pnictogen (As, Sb, or Bi) \cite{Makovicky_2006}, with the pure tennantite itself represented by the formula Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13}.
Despite being known for over two centuries, these minerals are still under investigation, both due to  structural peculiarities and potential applications. Arsenic- and antimony-containing compounds of this class are the advanced thermoelectric materials with a low lattice thermal conductivity\cite{Sootsman2009,Chetty2015}. 
\textcolor{red}{For example, the studies of tetrahedrite doped with Mn, Fe, Co, Ni, Zn, Te, Se showed ZT values in the range from 0.2 to 1.13\cite{Heo2014,Suekuni2013,Lu2012,Barbier2016,Rout2020,Zhu2019}.}

\textcolor{red}{The tennantite structure treated as sphalerite-type and belongs to \hmn{I -4 3 m} space group\cite{yaroslavzev2019,Makovicky_2006}.
There are three copper position: Cu1 is tetrahedrally coordinated by four sulpfur atoms, Cu2 is coordinated by three sulpfur atoms and Cu21 is tetrahedrally coordinated by three sulpfur and arsenic atoms. 
The one arsenic is coordinated by three sulpfur atoms.
Cu21 and Cu2 atoms possess very large atomic displacements perpendicular to the plane of the triangle (Fig.~\ref{fig:generel_view} right).}
In a case of tennantite- and tetrahedrite-based materials, it is generally considered, that a low lattice thermal conductivity is caused by hybridized Cu {\it 3d}  and chalcogen \textcolor{red}{{\it p}} orbitals\cite{Lai2015}.
Such a hybridization leads to a bonding asymmetry, and out-of-plane low frequency rattling modes, which are quasi-localized and anharmonic\cite{May2016,Bouyrie2015}.

@@

Some authors note\cite{DiBenedetto2005,Tanaka2015}, that asymmetric bonds give rise to magnetic properties and, in some cases, to antiferromagnetic (AFM) ordering\cite{yaroslavzev2019,Blandy_2018}.
In both kinds of materials, deviations from the Curie-Weiss law in the temperature dependencies of the magnetization were revealed: at 124~K\cite{Tanaka2015} in tennantite, and at 84~K\cite{Nasonova2016_2} in tetrahedrite.

@@

Large static values for Cu2 atom (Table~\ref{tab:enst_temp}) indicate the potential instability of this position. 
\textcolor{red}{According to structurel study of tennantie Cu2\cite{Hathwar2019} position may be describied as 6 differnt Cu position.
But according our study even after dividing Cu2 position into Cu2 and Cu 21, the Cu2 position remains highly disordered and there are no reasons for adding additional positions.}
This disordering is reliably described by the anharmonic model of atomic displacements using the Gram-Charlier expansion to the 4th order.

@@

Within \textcolor{red}{approach above}, the Einstein solid model is used to describe independent harmonic oscillators.
If so, the low energy spectrum should consist of peaks with the Einstein's energies, which we found by DebyeFit (Table~\ref{tab:enst_temp}).
Results of calculations are presented in Figure~\ref{fig:low_energ_raman}.
The Einstein's energies from structural study were used as the initial points for the fit, and in Table~\ref{tab:ram_fit} the initial  and fitted wavenumbers are presented.
The low energy $l_{1}$ and $l_{2}$ peaks are well described by Einstein's energies of Cu2, Cu21, Cu1, and As (66, 40, 79, and 122 cm\textsuperscript{-1} respectively).
\textcolor{red}{Same modes were observed and analized in tetrahedrites\cite{May2016,Lai2015} with energies approximately 4, 9 and 18 meV (33, 72, 145 cm\textsuperscript{-1}). 
The enegries of tetragidrite linked to the trigonally-bonded Cu atoms: the lowest energy is linked to Cu2 (they do not divide to Cu2 and Cu21), the middle --- to Cu1 and the highest --- to Sb. That in good agreement with our results.
Thus, the reasons for the strong anharmonic behavior, described in the papers, is a dynamic disorder of the copper atoms and arsenic atoms.}
Consequently, this leads to out-of-plane low frequency rattling modes, and a low lattice thermal conductivity.

@@

\section{Conclusions}\label{sec:level1}
In conclusion, the structural variety of synthetic tennantite, Cu\textsubscript{12}As\textsubscript{4}S\textsubscript{13},  leads to  a dynamic disorder of the copper atoms, asymmetric bonds, and \textcolor{red}{different charge distribution}.
This brings to the out-of-plane low frequency rattling modes, well described by Einstein characteristic temperatures: 74, 104, 115, and 185~K.
The peculiarities of the tennantite crystal and electronic structures lead to asymmetric Fermi surface, thus makinkg tennantite--tetrahedrite group compounds attractive research objects.
